---
title: "Drawing plasmid maps in R with plasmapR"
author: "Sara Burgess"
date: "2026-10-10"
format: html
editor: visual
execute:
  warning: false
  message: false
---

In this blog I'll run through how I drew plasmid maps in R using the plasmapR package (written by Brady Johnston). A plasmid map shows the genes on a circular plasmid as arrows around a ring, so you can see at a glance where the antimicrobial resistance (AMR) genes sit and what is around them. plasmapR does most of the work, but I wanted to change the colours, the labels, and which genes were shown, so a fair bit of this post is about modifying the ggplot object that plasmapR returns. I'll start by drawing a map for a single plasmid.

The plasmid sequence originates from an *Escherichia coli* strain isolated from dairy cattle faeces The plasmid was assembled using Flye and annotated using Prokka as described in [Burgess et al (2021)](https://doi.org/10.1016/j.idairyj.2021.104998).

## Packages

plasmapR isn't on CRAN, so it needs to be installed from GitHub.

```{r}
#| eval: false

devtools::install_github("bradyajohnston/plasmapr")
```

Along with plasmapR and the tidyverse, I used patchwork (Thomas Lin Pedersen) to combine the maps, and paletteer and RColorBrewer to choose the colours.

```{r}
library(tidyverse)
library(plasmapR)
library(paletteer)
library(patchwork)
library(RColorBrewer)
```

### Reading in the data

`read_gb()` reads the GenBank file into a plasmid object. I also saved the contig name (the file name without the folder or the `.gb` extension).

```{r}
fl <- "2026/2026-10-10-plasmidmap/data/pEE072d2_CTX-M.gb"
fl <- "2026/2026-10-10-plasmidmap/data/contig_3_v2.gb"
fl_name <- fl |>
  str_remove(pattern = "\\.gb$") |>
  str_remove("2026/2026-10-10-plasmidmap/data/")
plasmid <- fl |> read_gb()
```

plasmapR can convert the plasmid object straight into a data frame with `as.data.frame()`. 

```{r}
df <- plasmid |>
  as.data.frame()
```

Here is the first six rows of this dataframe:

```{r}
head(df)
```

### Grouping the genes by predicted function

This dataframe contains three rows that aren't coding sequences. The source:

```{r}
df |> filter(type == "source")
```

And the miscellenous sequences:

```{r}
df |> filter(type == "misc_feature")
```


I kept the coding sequences (CDS) and the miscellaneous features, which drops the source row but keeps the origin of replication (Ori) and the IS903 insertion sequence. I then used `mutate()` and `case_when()` to add a column grouping each gene by its predicted function, based on the gene name. `case_when()` works through the conditions in order and stops at the first match, so the order matters here. For example, *eitA* is named "eitA (Yi 2010) hypothetical protein CDS", so the iron uptake check (*eitA*-*eitD*) has to come before the hypothetical protein check. The transfer genes are picked up by the start of their names (*tra*, *trb* and *finO*), and the CTX-M gene is kept as its own ESBL/AmpC group. Genes with only a locus tag (e.g. EE0072d_05113) are grouped with the hypothetical proteins. Anything not matched ends up as "Other".

The CTX-M gene is named *bla_1* in the annotation, so I renamed it to *blaCTX-M* for the map label. Finally, I selected the columns `plot_plasmid()` expects, with the new function groups becoming the `type` column.

```{r}
df1 <- df |>
  filter(type %in% c("CDS", "misc_feature")) |>
  mutate(
    annotation = case_when(
      str_detect(name, "IS903") ~ "Plasmid mobilisation",
      str_detect(name, "^bla") ~ "ESBL/AmpC",
      str_detect(name, "^Ori$|^rep") ~ "Plasmid replication",
      str_detect(name, "^par") ~ "Plasmid partitioning",
      str_detect(name, "^relE|^pndA") ~ "Toxin-antitoxin",
      str_detect(name, "^tra|^trb|^finO") ~ "Plasmid transfer",
      str_detect(name, "^eit") ~ "Iron uptake",
      str_detect(name, "^EE0072d_|hypothetical") ~ "Hypothetical protein",
      TRUE ~ "Other"
    ),
    name = if_else(str_detect(name, "^bla"), "blaCTX-M", name)
  ) |>
  select(index, name, type = annotation, start, end, direction)
```

I didn't want the hypothetical proteins or the "Other" genes on the map, so I filtered them out. I also added a `mid` column (the midpoint of each gene), which I'll use later to position the gene labels.

```{r}
filtered_df <- df1 |>
  filter(
    type %in%
      c(
        "Plasmid replication",
        "Plasmid transfer",
        "Plasmid partitioning",
        "Plasmid mobilisation",
        "ESBL/AmpC",
        "Toxin-antitoxin",
        "Iron uptake"
      )
  ) |>
  mutate(mid = (start + end) / 2)
```

### Choosing the colours

The colours are mostly from RColorBrewer's "Paired" palette. I gave related groups the same colour (replication and partitioning share a light blue), used orange for the ESBL/AmpC gene so it stands out, green for the iron uptake genes, and a grey for the mobilisation genes so they sit in the background.

```{r}
custom_colors <- c(
  "Plasmid replication" = "#A6CEE3",
  "Plasmid transfer" = "#1F78B4",
  "Plasmid mobilisation" = "#B3B7B8FF",
  "Plasmid partitioning" = "#A6CEE3",
  "ESBL/AmpC" = "#FF7F00",
  "Toxin-antitoxin" = "#6A3D9A",
  "Iron uptake" = "#33A02C"
)
```

### Building the plot

`plot_plasmid()` returns a ggplot object, which means it can be changed like any other ggplot. Here's the default map.

```{r}
pl1 <- filtered_df |> plot_plasmid()
pl1
```

The default map labels every gene, which gets crowded quickly. The plot is built from several layers, and layers 2, 4, and 5 are the `geom_label_repel()` labels, the `geom_fit_text()` text, and the `annotate()` layer. I removed these three layers from the plot.

```{r}
pl1$layers <- pl1$layers[-c(2, 4, 5)]
```

I only wanted to label the resistance genes, so I made a vector of the function groups that shouldn't be labelled. I then added my own `geom_text()` layer using only the genes not in that list, positioned at the midpoint of each gene and nudged out from the ring. `append(..., after = 0)` adds the new text layer as the first layer, so it is drawn underneath the gene arrows. Setting `inherit.aes = FALSE` stops the text layer picking up the aesthetics from the plasmapR layers.

```{r}
no_text <- c(
  "Plasmid mobilisation",
  "Plasmid transfer",
  "Plasmid partitioning",
  "Toxin-antitoxin",
  "Plasmid replication",
  "Iron uptake"
)
pl1_modified <- pl1
pl1_modified$layers <- append(
  pl1_modified$layers,
  geom_text(
    data = filtered_df |>
      filter(!type %in% no_text),
    aes(x = mid, y = 3, label = name),
    inherit.aes = FALSE,
    size = 3,
    color = "black",
    nudge_y = 0.5,
    show.legend = FALSE
  ),
  after = 0
)
```

Finally, I added the custom colours, removed the background and axes with `theme_void()`, and renamed the legend.

```{r}
max_pos <- max(filtered_df$end)
custom_plot <- pl1_modified +
  scale_fill_manual(values = custom_colors) +
  theme_void() +
  labs(fill = "Predicted function")

custom_plot
```

<-- Add something here generally about using a function and patchwork for multiple maps -->

## Saving the figure

Finally, I saved the combined figure as an SVG, a PDF, and a PNG. The SVG was saved at 190 x 282.9 mm, while the PDF and PNG were saved at 200 x 150 mm (a width of 200 mm multiplied by `fig_ratio`, 12/16, for the height).

```{r}
#| eval: false

fig_ratio <- 12 / 16
height_ratio <- 282.9 / (180 * fig_ratio)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.svg",
  dpi = 1000,
  width = 190,
  height = 282.9,
  units = "mm"
)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.pdf",
  device = "pdf",
  dpi = 1000,
  width = 200,
  height = 200 * fig_ratio,
  units = "mm"
)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.png",
  device = "png",
  dpi = 1000,
  width = 200,
  height = 200 * fig_ratio,
  units = "mm"
)
```

## Key tips

- `plot_plasmid()` returns a ggplot object, so you can remove, reorder, or add layers (`pl1$layers`) and change layer parameters directly when the defaults don't quite fit.
- If some of the fields you need are nested, `map_dfr()` with `%||% NA` is a simple way to flatten them into a data frame.
- The order of the conditions in `case_when()` matters. Put the most specific groups first.
- Get the map right for one plasmid before wrapping the code in a function. Returning a list from the function means you can pull out other information (like the AMR gene counts) at the same time as the plots.
- Much of the work happens before R. Check your annotations (e.g. against the ResFinder output) before you draw the map.