---
title: "Drawing plasmid maps in R with plasmapR"
author: "Sara Burgess"
date: "2026-10-10"
format: html
editor: visual
execute:
  warning: false
  message: false
---

In this blog I'll run through how I drew plasmid maps in R using the plasmapR package (written by Brady Johnston). I wanted to change the colours, the labels, and which genes were shown, so I modified the ggplot object that plasmapR returns. I'll be drawing a map for a single plasmid. You can download the plasmid sequence [here](https://github.com/sburgess2/blogs/tree/main/2026/2026-10-10-plasmidmap/data).

The plasmid sequence originates from an *Escherichia coli* strain isolated from dairy cattle faeces. The plasmid was assembled using Flye and annotated using Prokka as described in [Burgess et al (2021)](https://doi.org/10.1016/j.idairyj.2021.104998).

## Packages

plasmapR isn't on CRAN, so it needs to be installed from GitHub.

```{r}
#| eval: false

devtools::install_github("bradyajohnston/plasmapr")
```

First load the libraries we'll be using.

```{r}
library(tidyverse)
library(plasmapR)
library(paletteer)
library(RColorBrewer)
```

### Reading in the data

Next read in the GenBank file into the plasmid object with `read_gb()`.

```{r}
#fl <- "2026/2026-10-10-plasmidmap/data/pEE072d2_CTX-M.gb"
fl <- "2026/2026-10-10-plasmidmap/data/contig_3_v2.gb"
plasmid <- fl |> read_gb()
```

Then we can convert the plasmid object into a data frame using `as.data.frame()`.

```{r}
df <- plasmid |>
  as.data.frame()
```

Here are the first six rows of this data frame:

```{r}
head(df)
```

### Grouping the genes by predicted function

This data frame three types of sequences. The source:

```{r}
df |> filter(type == "source")
```

The miscellaneous sequences:

```{r}
df |> filter(type == "misc_feature")
```

The miscellaneous sequences:

```{r}
df |> filter(type == "cds")
```

I wanted to keep the origin of replication (Ori) and the IS903 insertion sequence on the map, so I kept the miscellaneous features as well as the coding sequences (CDS). The following only drops the source row.

```{r}
df |>
  filter_out(type == "source"))
```

Next, we'll use `mutate()` and `case_when()` to add a column called `annotation`, which will group each gene by its predicted function (Plasmid mobilisation, Plasmid replication and partitioning, Plasmid transfer, Antimicrobial resistance, and Other), based on the gene name. `case_when()` works through the conditions in order and stops at the first match, so the order matters. The `str_detect` function is used to look for a pattern (in this case certain gene names) in a string (in this case the `name` column). Anything not matched, including the hypothetical proteins (e.g. EE0072d_05113), ends up as "Other".

```{r}
df |>
  filter(type %in% c("CDS", "misc_feature")) |>
  mutate(
    annotation = case_when(
      str_detect(name, "IS903") ~ "Mobilisation",
      str_detect(name, "^bla") ~ "Antimicrobial resistance",
      str_detect(
        name,
        "^Ori$|^rep|^par"
      ) ~ "Plasmid replication and partitioning",
      str_detect(name, "^tra|^trb|^finO") ~ "Plasmid transfer",
      TRUE ~ "Other"
    ),
    name = str_replace(name, "bla_1", "blaCTX-M")
  )
```

The *bla*~CTX-M~ gene is named *bla_1* in the annotation, so in the last line above I renamed it to *blaCTX-M* for the map label.

Finally, we need to select the columns `plot_plasmid()` expects, with the new function groups becoming the `type` column.

```{r}
df |>
  select(index, name, type = annotation, start, end, direction)
```

I wanted all the genes to be visible on the map, so I kept the "Other" genes too. We also need to add a `mid` column (the midpoint of each gene), which is used later to position the gene labels.

```{r}
df |>
  mutate(mid = (start + end) / 2)
```

We can then chain the code above together and save it into a new object df1
```{r}
df1 <- df |>
  filter(type %in% c("CDS", "misc_feature")) |>
  mutate(
    annotation = case_when(
      str_detect(name, "IS903") ~ "Plasmid mobilisation",
      str_detect(name, "^bla") ~ "ESBL/AmpC",
      str_detect(
        name,
        "^Ori$|^rep|^par"
      ) ~ "Plasmid replication and partitioning",
      str_detect(name, "^relE|^pndA") ~ "Toxin-antitoxin",
      str_detect(name, "^tra|^trb|^finO") ~ "Plasmid transfer",
      str_detect(name, "^eit") ~ "Iron uptake",
      TRUE ~ "Other"
    ),
    name = str_replace(name, "bla_1", "blaCTX-M")
  ) |>
  select(index, name, type = annotation, start, end, direction) |>
  mutate(mid = (start + end) / 2)
```

If you don't want all the genes to be displayed on the plasmid map you could filter them using:

```{r}
```

### Choosing the colours

Where I can I like to select colours that match/relevant to my topic. Given this plasmid contains an antimicrobial resistance gene I decided to use some of the colours used for AMR awareness week from these two images.

I uploaded them into https://imagecolorpicker.com/ to get the hex codes for the two shades of blue and the black from the image and the orange from the second image and a light grey for the "Other" genes so they sit in the background.

We'll use these colours to create a custom palette as follows:

```{r}
custom_colors <- c(
  "Plasmid replication and partitioning" = "#cbe8f6",
  "Plasmid transfer" = "#1898d5",
  "Mobilisation" = "#010103",
  "ESBL" = "#fe6f43",
  "Other" = "#D9D9D9"
)
```

### Building the plot

`plot_plasmid()` returns a ggplot object. Here's the default map.

```{r}
pl1 <- df1 |> plot_plasmid()
pl1
```

The default map labels every gene, which gets crowded. After a lot of playting around I realised that removing some of the layers was the key to getting the map into a format I liked more. The plot is built from layers, which are stored as a list in `pl1$layers`. 

```{r}
pl1$layers
```

There are five layers. Layer 1 draws the plasmid backbone (the ring), and layer 3 draws the gene arrows. These are the two layers I kept. Layers 2, 4, and 5 are the `geom_label_repel()` labels, the `geom_fit_text()` text, and the `annotate()` layer. To see what each layer adds, we'll remove them one at a time. Removing items from a list by position works the same way as removing items from a vector, using a minus sign in front of the position.

Removing only layer 1 drops the plasmid backbone, leaving the gene arrows and labels without the ring.

```{r}
pl_no_backbone <- pl1
pl_no_backbone$layers <- pl_no_backbone$layers[-1]
pl_no_backbone
```

Removing only layer 2 drops the `geom_label_repel()` labels, which sit outside the ring with lines pointing to each gene.

```{r}
pl_no_repel <- pl1
pl_no_repel$layers <- pl_no_repel$layers[-2]
pl_no_repel
```

Removing only layer 3 drops the gene arrows, leaving the ring and labels without the genes.

```{r}
pl_no_arrows <- pl1
pl_no_arrows$layers <- pl_no_arrows$layers[-3]
pl_no_arrows
```

Removing only layer 4 drops the `geom_fit_text()` text, which fits the gene names inside the arrows.

```{r}
pl_no_fit_text <- pl1
pl_no_fit_text$layers <- pl_no_fit_text$layers[-4]
pl_no_fit_text
```

Removing only layer 5 drops the `annotate()` layer, which adds text to the centre of the map.

```{r}
pl_no_annotate <- pl1
pl_no_annotate$layers <- pl_no_annotate$layers[-5]
pl_no_annotate
```

Now that we know what each layer does, we can remove all three from `pl1`.

```{r}
pl1$layers <- pl1$layers[-c(2, 4, 5)]
```

I didn't want the "Other" genes labelled. To do this, we'll add our own `geom_text()` layer using only the genes that aren't in the "Other" group, positioned at the midpoint of each gene and nudged out from the ring. `append(..., after = 0)` adds the new text layer as the first layer, so it is drawn underneath the gene arrows. Setting `inherit.aes = FALSE` stops the text layer picking up the aesthetics from the plasmapR layers.

```{r}
pl1_modified <- pl1
pl1_modified$layers <- append(
  pl1_modified$layers,
  geom_text(
    data = df1 |>
      filter(type != "Other"),
    aes(x = mid, y = 3, label = name),
    inherit.aes = FALSE,
    size = 3,
    color = "black",
    nudge_y = 0.5,
    show.legend = FALSE
  ),
  after = 0
)
```

Finally, we'll add the custom colours, remove the background and axes with `theme_void()`, and rename the legend.

```{r}
max_pos <- max(filtered_df$end)
custom_plot <- pl1_modified +
  scale_fill_manual(values = custom_colors) +
  theme_void() +
  labs(fill = "Predicted function")

custom_plot
```

\<-- Add something here generally about using a function and patchwork for multiple maps --\>

## Saving the figure

Finally, I saved the combined figure as an SVG, a PDF, and a PNG. The SVG was saved at 190 x 282.9 mm, while the PDF and PNG were saved at 200 x 150 mm (a width of 200 mm multiplied by `fig_ratio`, 12/16, for the height).

```{r}
#| eval: false

fig_ratio <- 12 / 16
height_ratio <- 282.9 / (180 * fig_ratio)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.svg",
  dpi = 1000,
  width = 190,
  height = 282.9,
  units = "mm"
)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.pdf",
  device = "pdf",
  dpi = 1000,
  width = 200,
  height = 200 * fig_ratio,
  units = "mm"
)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.png",
  device = "png",
  dpi = 1000,
  width = 200,
  height = 200 * fig_ratio,
  units = "mm"
)
```

## Key tips

- `plot_plasmid()` returns a ggplot object, so you can remove, reorder, or add layers (`pl1$layers`) and change layer parameters directly when the defaults don't quite fit.
- The order of the conditions in `case_when()` matters. Put the most specific groups first.
- Get the map right for one plasmid before wrapping the code in a function. Returning a list from the function means you can pull out other information (like the AMR gene counts) at the same time as the plots.
- Much of the work happens before R. Check your annotations (e.g. against the ResFinder output) before you draw the map.