---
title: "Drawing plasmid maps in R with plasmapR"
author: "Sara Burgess"
date: "2026-10-10"
format: html
editor: visual
execute:
  warning: false
  message: false
---

In this blog I'll run through how I drew plasmid maps in R using the plasmapR package (written by Brady Johnston). I wanted to change the colours, the labels, and which genes were shown, so I modified the ggplot object that plasmapR returns. I'll be drawing a map for a single plasmid. You can download the plasmid sequence [here](https://github.com/sburgess2/blogs/tree/main/2026/2026-10-10-plasmidmap/data).

The plasmid sequence originates from an *Escherichia coli* strain isolated from dairy cattle faeces. The plasmid was assembled using Flye and annotated using Prokka as described in [Burgess et al (2021)](https://doi.org/10.1016/j.idairyj.2021.104998).

## Packages

plasmapR isn't on CRAN, so it needs to be installed from GitHub.

```{r}
#| eval: false

devtools::install_github("bradyajohnston/plasmapr")
```

First load the libraries we'll be using.

```{r}
library(tidyverse)
library(plasmapR)
library(paletteer)
library(RColorBrewer)
```

### Reading in the data

Next read in the GenBank file into the plasmid object with `read_gb()`.

```{r}
#fl <- "2026/2026-10-10-plasmidmap/data/pEE072d2_CTX-M.gb"
fl <- "2026/2026-10-10-plasmidmap/data/contig_3_v2.gb"
plasmid <- fl |> read_gb()
```

Then we can convert the plasmid object into a data frame using `as.data.frame()`.

```{r}
df <- plasmid |>
  as.data.frame()
```

Here are the first six rows of this data frame:

```{r}
head(df)
```

### Grouping the genes by predicted function

This data frame three types of sequences. The source:

```{r}
df |> filter(type == "source")
```

The miscellaneous sequences:

```{r}
df |> filter(type == "misc_feature")
```

The miscellaneous sequences:

```{r}
df |> filter(type == "cds")
```

I wanted to keep the origin of replication (Ori) and the IS903 insertion sequence on the map, so I kept the miscellaneous features as well as the coding sequences (CDS). The following only drops the source row.

```{r}
df |>
  filter_out(type == "source"))
```

Next, we'll use `mutate()` and `case_when()` to add a column called `annotation`, which will group each gene by its predicted function (Plasmid mobilisation, Plasmid replication and partitioning, Plasmid transfer, Antimicrobial resistance, and Other), based on the gene name. `case_when()` works through the conditions in order and stops at the first match, so the order matters. The `str_detect` function is used to look for a pattern (in this case certain gene names) in a string (in this case the `name` column). Anything not matched, including the hypothetical proteins (e.g. EE0072d_05113), ends up as "Other". The last line renames ...

```{r}
df |>
  filter(type %in% c("CDS", "misc_feature")) |>
  mutate(
    annotation = case_when(
      str_detect(name, "IS903") ~ "Plasmid mobilisation",
      str_detect(name, "^bla") ~ "Antimicrobial resistance",
      str_detect(
        name,
        "^Ori$|^rep|^par"
      ) ~ "Plasmid replication and partitioning",
      str_detect(name, "^tra|^trb|^finO") ~ "Plasmid transfer",
      TRUE ~ "Other"
    ),
    name = if_else(str_detect(name, "^bla"), "blaCTX-M", name)
  )
```

The CTX-M gene is named *bla_1* in the annotation, so I renamed it to *blaCTX-M* for the map label. Finally, we need to select the columns `plot_plasmid()` expects, with the new function groups becoming the `type` column.

```{r}
df1 <- df |>
  filter(type %in% c("CDS", "misc_feature")) |>
  mutate(
    annotation = case_when(
      str_detect(name, "IS903") ~ "Plasmid mobilisation",
      str_detect(name, "^bla") ~ "ESBL/AmpC",
      str_detect(
        name,
        "^Ori$|^rep|^par"
      ) ~ "Plasmid replication and partitioning",
      str_detect(name, "^relE|^pndA") ~ "Toxin-antitoxin",
      str_detect(name, "^tra|^trb|^finO") ~ "Plasmid transfer",
      str_detect(name, "^eit") ~ "Iron uptake",
      TRUE ~ "Other"
    ),
    name = if_else(str_detect(name, "^bla"), "blaCTX-M", name)
  ) |>
  select(index, name, type = annotation, start, end, direction)
```

I wanted all the genes to be visible on the map, so I kept the "Other" genes too. We also need to add a `mid` column (the midpoint of each gene), which is used later to position the gene labels.

```{r}
filtered_df <- df1 |>
  mutate(mid = (start + end) / 2)
```

### Choosing the colours

The colours are mostly from RColorBrewer's "Paired" palette. I used orange for the ESBL/AmpC gene so it stands out, green for the iron uptake genes, black for the mobilisation genes, and a light grey for the "Other" genes so they sit in the background.

```{r}
custom_colors <- c(
  "Plasmid replication and partitioning" = "#A6CEE3",
  "Plasmid transfer" = "#1F78B4",
  "Plasmid mobilisation" = "black",
  "ESBL/AmpC" = "#FF7F00",
  "Toxin-antitoxin" = "#6A3D9A",
  "Iron uptake" = "#33A02C",
  "Other" = "#D9D9D9"
)
```

### Building the plot

`plot_plasmid()` returns a ggplot object, which means it can be changed like any other ggplot. Here's the default map.

```{r}
pl1 <- filtered_df |> plot_plasmid()
pl1
```

The default map labels every gene, which gets crowded. Changing the layers was the key step in getting the map into a format I liked more. Like any ggplot, the plot is built from layers, which are stored as a list in `pl1$layers`. We can print this list to see which geom each layer uses.

```{r}
pl1$layers
```

There are five layers. Layer 1 draws the plasmid backbone (the ring), and layer 3 draws the gene arrows. These are the two layers I kept. Layers 2, 4, and 5 are the `geom_label_repel()` labels, the `geom_fit_text()` text, and the `annotate()` layer. To see what each of these layers adds, we'll remove them one at a time. Removing items from a list by position works the same way as removing items from a vector, using a minus sign in front of the position.

Removing layer 2 drops the `geom_label_repel()` labels, which sit outside the ring with lines pointing to each gene.

```{r}
pl_step1 <- pl1
pl_step1$layers <- pl_step1$layers[-2]
pl_step1
```

Removing layer 4 as well drops the `geom_fit_text()` text, which fits the gene names inside the arrows.

```{r}
pl_step2 <- pl1
pl_step2$layers <- pl_step2$layers[-c(2, 4)]
pl_step2
```

Removing layer 5 drops the `annotate()` layer, which adds text to the centre of the map.

```{r}
pl_step3 <- pl1
pl_step3$layers <- pl_step3$layers[-c(2, 4, 5)]
pl_step3
```

Each step starts from a copy of `pl1`, so the layer numbers always match the original plot. If we removed layer 2 first and then tried to remove layer 4 from the shorter list, we would remove the wrong layer. Now that we know what each layer does, we can remove all three from `pl1`.

```{r}
pl1$layers <- pl1$layers[-c(2, 4, 5)]
```

I only wanted to label the resistance genes. To do this, we'll make a vector of the function groups that shouldn't be labelled, then add our own `geom_text()` layer using only the genes not in that list, positioned at the midpoint of each gene and nudged out from the ring. `append(..., after = 0)` adds the new text layer as the first layer, so it is drawn underneath the gene arrows. Setting `inherit.aes = FALSE` stops the text layer picking up the aesthetics from the plasmapR layers.

```{r}
no_text <- c(
  "Plasmid mobilisation",
  "Plasmid transfer",
  "Plasmid replication and partitioning",
  "Toxin-antitoxin",
  "Iron uptake",
  "Other"
)
pl1_modified <- pl1
pl1_modified$layers <- append(
  pl1_modified$layers,
  geom_text(
    data = filtered_df |>
      filter(!type %in% no_text),
    aes(x = mid, y = 3, label = name),
    inherit.aes = FALSE,
    size = 3,
    color = "black",
    nudge_y = 0.5,
    show.legend = FALSE
  ),
  after = 0
)
```

Finally, we'll add the custom colours, remove the background and axes with `theme_void()`, and rename the legend.

```{r}
max_pos <- max(filtered_df$end)
custom_plot <- pl1_modified +
  scale_fill_manual(values = custom_colors) +
  theme_void() +
  labs(fill = "Predicted function")

custom_plot
```

\<-- Add something here generally about using a function and patchwork for multiple maps --\>

## Saving the figure

Finally, I saved the combined figure as an SVG, a PDF, and a PNG. The SVG was saved at 190 x 282.9 mm, while the PDF and PNG were saved at 200 x 150 mm (a width of 200 mm multiplied by `fig_ratio`, 12/16, for the height).

```{r}
#| eval: false

fig_ratio <- 12 / 16
height_ratio <- 282.9 / (180 * fig_ratio)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.svg",
  dpi = 1000,
  width = 190,
  height = 282.9,
  units = "mm"
)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.pdf",
  device = "pdf",
  dpi = 1000,
  width = 200,
  height = 200 * fig_ratio,
  units = "mm"
)

ggsave(
  plot = combined_plot,
  filename = "figures/plasmid_map.png",
  device = "png",
  dpi = 1000,
  width = 200,
  height = 200 * fig_ratio,
  units = "mm"
)
```

## Key tips

- `plot_plasmid()` returns a ggplot object, so you can remove, reorder, or add layers (`pl1$layers`) and change layer parameters directly when the defaults don't quite fit.
- The order of the conditions in `case_when()` matters. Put the most specific groups first.
- Get the map right for one plasmid before wrapping the code in a function. Returning a list from the function means you can pull out other information (like the AMR gene counts) at the same time as the plots.
- Much of the work happens before R. Check your annotations (e.g. against the ResFinder output) before you draw the map.