---
title: "Demystifying ggplot2"
author: "Sara Burgess"
date: last-modified
format: html
---

Recently I was speaking to a PhD student about R. Our conversation went something like this (I'm paraphrasing):

> Student: "I hate R."
>
> Me: "Is it the modelling part you don't like?"
>
> Student: "All of it. R and I just don't see eye to eye."
>
> Me: "What about generating plots in R using ggplot2?"
>
> Student: "No, that takes so long, particularly adding in colours. I'd love to be able to generate beautiful plots though."
>
> Me: "It took me quite a while to understand how to use a custom palette and it's also quite fiddly getting the hang of all the different options you can add into the theme function to format a ggplot2 plot."

Hopefully you get the idea that she really doesn't like R. I felt for her, as writing up a PhD is hard enough without adding something else on top. This blog is for her, and for anyone else out there who feels like they'll never get their head around R and ggplot2. This was me a few years ago. I used to get frustrated with R and thought I'd never get the hang of it. I only started to get my head around it through practice, and it wasn't until ChatGPT emerged that I stopped feeling so frustrated. However, despite AI being a huge help with coding, it still helps to understand the basics and how to build up a plot. It wasn't until I began reading Nicola Rennie's blogs that I really appreciated the intricacies of plot design using ggplot2. It's not just about producing a beautiful plot; it's also about generating one that's easy for your audience to understand. In this blog I'll walk through how I build up a plot using ggplot2.

```{r}
#| echo: false
#| message: false
#| warning: false

#load libraries
#library(tidyverse)
library(gt)
#library(ggtext)
```

First load the libraries that we'll be using :

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(ggtext)

```

In this example we'll be generating a heatmap using a simulated dataset. This dataset gives the number of contigs carrying each antimicrobial resistance gene, broken down by bacterial family. The numbers are made up. You can load the data from here:

```{r}
#| message: false
#| warning: false
#| eval: false

df <- read_csv("https://raw.githubusercontent.com/sburgess2/blogs/main/2026/2026-08-05-demystifying-ggplot2/data/arg_contigs.csv")
```

This is what the first six lines of this dataframe looks like:

```{r}

head(df)

```

```{r}
#| echo: false
#| message: false
#| warning: false
df <- read_csv("data/arg_contigs.csv")
```

The first step is to tell R/ggplot2 what data you're using. Here, within the `ggplot()` function, we're telling R to use the data frame called `df`, that `gene` goes on the x-axis, `family` goes on the y-axis, and that `fill` specifies each tile will be coloured by `n_contigs`. When you run this you'll see the plot panel with the axes drawn but no data plotted.

```{r}
#| message: false
#| warning: false

ggplot(data = df, aes(x = gene, y = family, fill = n_contigs))
```

To actually plot the data we need to tell R what type of plot we want. We do this by adding a geom layer with `+`. There are a whole family of geoms, all starting with `geom_`, and if you start typing that in your editor you'll get a list of options. There are a lot of different geoms, plus many extensions you can use with ggplot2 (you can find a list [here](https://exts.ggplot2.tidyverse.org/gallery/)). We're generating a heatmap, so we'll use `geom_tile()`. Other packages provide their own heatmap functions too. Check out these blogs [here](https://r-graph-gallery.com/215-the-heatmap-function.html) and [here](https://www.datanovia.com/learn/machine-learning/clustering/heatmap) if you want to learn more about other heatmap functions.

```{r}
#| message: false
#| warning: false

ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile(show.legend = FALSE)
```

We now have a basic heatmap. Before we continue, let's have a look at some of the aesthetics you can use within `geom_tile()`. You can find out more about how to use a function and its arguments by typing `?` followed by the function name, for example `?geom_tile`.

Generally, ggplot2 geoms share the same set of core arguments e.g. `mapping`, `data`, `show.legend`, `inherit.aes` and a few others. These are always listed in the help file (but sometimes they're tricky to follow). The `mapping` and `data` arguments are only needed if a layer is doing something different from the rest of the plot; otherwise they're specified once in `ggplot()`. Generally for a `geom_tile()` you won't need to change the defaults for these core arguments. However, you may want to use `inherit.aes = FALSE` if you don't want to inherit the plot data from `ggplot()` or use `show.legend = FALSE` if you do not want a legend.

The things that you you are more likely to want to change in `geom_tile()` are passed through `...`, so they don't appear as named arguments, which is also why the help file can be confusing to read. For example, aesthetics such as those listed below.

```{r}
#| message: false
#| warning: false
#| echo: false

tribble(
  ~aesthetic, ~description,
  "fill", "Interior colour of the tile.",
  "colour", "Colour of the tile border. NA by default, so tiles are drawn without an outline.",
  "linewidth", "Thickness of the tile border. This has no visible effect until colour is set",
  "linetype", "Line style of the border, for example solid or dashed.",
  "alpha", "Opacity of the fill, from 0 (transparent) to 1 (opaque).",
  "width", "Width of each tile.",
  "height", "Height of each tile.") |>
     gt() |>
  cols_label(
    aesthetic = "Aesthetic",
    description = "Description"
  ) |>
  #tab_style(
   # style = cell_text(font = "monospace"),
    #locations = cells_body(columns = c(aesthetic, #description))
  #) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  cols_align(align = "left", columns = description) |> 
  tab_options(
    table.width = pct(100),
    table.font.size = px(14),
    data_row.padding = px(6),
    table.border.top.style = "none"
  )
```

Getting the hang of the different arguments is tricky. Part of it is practice, but part of it is accepting that you will need to look them up. I always do. Also ask yourself whether you really need to understand every argument. In this day and age of AI do you need to understand every line of code? Especially if coding isn't your thing.

For this heatmap we're going to use the colour and linewidth aesthetics.

```{r}
#| message: false
#| warning: false

p <- ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile(colour = "white", linewidth = 1)

p
```

We're now going to add layers to the plot. First, we'll add the axis and legend labels. That's all you need if it's for a publication, but if it's for a presentation or a blog I would also include a title and possibly a subtitle.

```{r}
#| message: false
#| warning: false

p_labels <- p +
    labs(x = "Antimicrobial resistance gene",
        y = "Family",
        fill = "Number of contigs"
    )

p_labels
```

The y-axis title, "Family" is currently rotated through 90 degrees and sitting halfway down the left-hand side. This is common practice for scientific publications but it requires the reader to tilt their head to read one word. I'd rather have it horizontal and above the y axis. There are arguments which allow the y axis title to be rotated and moved above axis within the theme function. You can try it using the following code:

```{r}
#| message: false
#| warning: false

p_labels +
    theme(
        axis.title.y = element_text(
            angle = 0,
            vjust = 1,
            hjust = 1
        )
    )
```

`angle = 0` rotates the title back to horizontal, and `vjust = 1` pushes it to the top of the axis. `hjust = 1` right-aligns it. However, because when the y-axis title is set using labs = it always sits to the left. I would prefer it to be positioned directly above the family names. One option is to use `labs(tag = )`. A tag can be positioned by coordinates (within the theme function), so you can drop it directly above the family names rather than out at the plot edge.

```{r}
#| message: false
#| warning: false

p_tag <- p_labels +
    labs(
        y = NULL,
        tag = "Family"
    ) +
    theme(
        plot.tag = element_text(hjust = 0),
        #size = rel(0.9)),
        plot.tag.position = c(0.02, 0.78),
        plot.tag.location = "plot"
    )

p_tag
```

`plot.tag.position` takes two numbers between 0 and 1, measured across the width and height of the plot. So `c(0.02, 0.78)` sits the tag 2% in from the left and 78% of the way up. `plot.tag.location = "plot"` measures those against the whole plot rather than the panel, and `hjust = 0` left-aligns the text at that point.

Both numbers need nudging by eye. Increase the first to shift the label right until it lines up over the family names, and adjust the second until it sits just above the top row. The values that work depend on the size of your plot, so if you change the figure dimensions you'll need to revisit them.

Worth knowing that a tag placed this way doesn't reserve any space for itself, it simply sits on top of whatever is underneath. That's what lets you tuck it into the gap, but it also means it will happily overlap your legend or the tiles if the coordinates are off.

Many figures in publications stop there, using the default formatting settings. However, to make your plot eye-catching and easy to read, choose colours wisely and declutter.

Let's change the colour palette to one that makes it easier to distinguish between the low and high values. You can either use a ready-made palette or create your own. The R Graph Gallery's [colour palette finder](https://r-graph-gallery.com/color-palette-finder) has a huge variety to choose from. If you find that list a bit overwhelming to look through, I'd recommend starting with the viridis palettes or the ColorBrewer palettes. For this example I'll use a ColorBrewer palette.

One thing to watch here: our tiles are coloured using `fill`, so we need a `scale_fill_*` function rather than `scale_colour_*`. Because `n_contigs` is a continuous variable we also need `scale_fill_distiller()`. `scale_fill_brewer()` looks like the obvious choice, but it only works with discrete variables. Within the `scale_fill_distiller()` argument we will use the arguments `palette` to specify which palette we want to use (you can find a list of them [here](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html)), and `direction` to specify the order where `direction = 1` is the default order of the palette and `direction = -1` reverses the order.

```{r}
#| message: false
#| warning: false

p_colour <- p_tag +
    scale_fill_distiller(palette = "YlGnBu", direction = 1)

p_colour
```

If you want to select the colours to use you can use either `scale_fill_gradient(), which` takes a low colour and a high colour and works out everything in between. If you want a midpoint colour as well, `scale_fill_gradient2()` gives you low, mid and high, which is useful for anything diverging around zero.

```{r}
#| message: false
#| warning: false
#| eval: false

p_tag +
    scale_fill_gradient(low = "#F7FEAE", high = "#045275")

p_tag +
    scale_fill_gradient2(low = "#F7FEAE", mid = "#7CCBA2", high = "#045275")
```

Custom palettes for categorical fills work differently. If you want to use a custom palette, you first create the palette, one colour per category, then add it into the `scale_fill_manual()` function:

```{r}
#| message: false
#| warning: false
#| eval: false

custom_colors <- c(
        "Category 1" = "#EDD746FF",      
        "Category 2" = "#00496FFF",         
        "Category 3" = "#DD4124FF", 
        "Category 4" = "#B3B7B8FF")

ggplot(df, aes(x = x, y = y)) +
geom_point() +
scale_fill_manual(values = custom_colors)
```

Back to our heatmap. You'll notice that where n_contig = 0 the tile was coloured light yellow (as per the palette). However, you may want a value of 0 to have no colour. There are multiple options. The first option is to set limits within the scale_fill_distiller() or `scale_fill_gradient()` function and assign white (or the background colour of your plot) to `na.value`.

```{r}
#| message: false
#| warning: false
#| eval: false



p_tag +
    scale_fill_distiller(
        palette = "YlGnBu",
        direction = 1,
        limits = c(1, 10),
        na.value = "white"
    )
```

Another option is to convert the zeros to `NA` first then set the argument `na.value` to white (the default colour is `grey50`). If your dataset already has missing values instead of the zeros, you only need to add the `na.value` argument.

```{r}
#| message: false
#| warning: false
#| eval: false

#Convert zeros to NA
df <- df |>
    mutate(n_contigs = na_if(n_contigs, 0))

p <- ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile(colour = "white")

p_labels <- p +
    labs(x = "Antimicrobial resistance gene",
        y = "Family",
        fill = "Number of contigs"
    )

p_labels +
    geom_tile(colour = "white", linewidth = 2) +
    scale_fill_distiller(
        palette = "YlGnBu",
        direction = 1,
        na.value = "white"
    )
```

Alternatively you can use a binned scale, in other words assign each value to a bin and use the `scale_fill_stepsn()` function.

```{r}
#| message: false
#| warning: false
#| eval: false

p_labels +
    scale_fill_stepsn(
        colours = c("#EDF8B1", "#7FCDBB", "#225EA8", "#253494"),
        breaks = c(2, 4, 6, 8),
        limits = c(1, 10),
        na.value = "white"
    )
```

You may have noticed when using ggplot2 that the grey panel background is a ggplot2 default. We can see the outline of it around the heatmap. To remove this grey background we will use `theme_minimal()`. There are many other themes that you could try out that are also quite clean e.g. `theme_light()` and `theme_classic()`

```{r}
#| message: false
#| warning: false

p_theme <- p_colour + theme_minimal()

p_theme
```

Everything else can be formatted within the `theme()` function. Before the theme itself I set the colours and fonts as objects at the top. It means changing the whole look of the plot is a single edit instead of fifteen, and if you're making a set of plots for the same report they'll match.

```{r}
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 7

text_col <- "#1A1A1A"
label_col <- "#4D4D4D"
font <- "Times New Roman"
bg_col <- "white"

p_final <- p_theme +
    theme_minimal(base_size = 12) +
    theme(
        legend.position = "top",
        legend.justification = "left",
        legend.location = "plot",
        legend.direction = "horizontal",
        legend.title = element_text(color = text_col, family = font),
        legend.text = element_text(color = label_col, family = font),
        plot.margin = margin(10, 10, 10, 10),
        axis.text = element_text(color = label_col, family = font, size = rel(0.8)),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        panel.background = element_rect(fill = bg_col, color = bg_col),
        plot.background = element_rect(fill = bg_col, color = bg_col),
        axis.ticks = element_blank()
    )

p_final
```

Working through what's in there.

The legend arguments put the key at the top, left aligned and horizontal. `legend.location = "plot"` aligns it to the plot edge.

`panel.grid = element_blank()` and `axis.ticks = element_blank()` remove marks a heatmap doesn't need. The tiles already show you where everything is.

`axis.text.x` is angled at 45 degrees. Generally I prefer my x axis labels to be horizontal but some of the gene names are quite long and won't easily fit horizontally, and angled labels are easier to read than vertical ones.

One last thing, the plot in the plot pane is drawn at whatever size the pane happens to be. Drag the pane wider and the text gets relatively smaller, make it narrower and your labels start colliding. You end up tuning text sizes against a moving target.

To view your plot in the same dimensions that it will be exported use the camcorder package `gg_record()` sets a fixed canvas, and from then on every plot you draw is rendered at those exact dimensions and shown in the viewer:

```{r}
#| eval: false

library(camcorder)

gg_record(
    dir = "output/recording",
    device = "png",
    width = 12,
    height = 7,
    units = "in",
    dpi = 300
)

ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile(colour = "white", linewidth = 1) +
    labs(
        x = "Antimicrobial resistance gene",
        y = NULL,
        fill = "Number of contigs",
        tag = "Family"
    ) +
    scale_fill_distiller(palette = "YlGnBu", direction = 1) +
    theme_minimal(base_size = 12) +
    theme(
        legend.position = "top",
        legend.justification = "left",
        legend.location = "plot",
        legend.direction = "horizontal",
        legend.title = element_text(color = text_col, family = font),
        legend.text = element_text(color = label_col, family = font),
        plot.tag = element_text(hjust = 0, color = text_col, family = font),
        plot.tag.position = c(0.02, 0.78),
        plot.tag.location = "plot",
        plot.margin = margin(10, 10, 10, 10),
        axis.text = element_text(color = label_col, family = font, size = rel(0.8)),
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        panel.background = element_rect(fill = bg_col, color = bg_col),
        plot.background = element_rect(fill = bg_col, color = bg_col),
        axis.ticks = element_blank()
    )

    gg_stop_recording()
```

To save the file use `ggsave()`:

```{r}
#| eval: false

ggsave(
    "output/arg_heatmap.png",
    p_final,
    width = 12,
    height = 7,
    dpi = 300,
    bg = bg_col
)
```

You me be asking the question that it would be a pain/nuisance writing out all the theme() code each time. There are a few options to avoid doing this. Write a template as described [here](https://nrennie.rbind.io/blog/script-templates-r/) by Nicola Rennie. Write a function for your preferred theme - describes one here. or if you are using Claude Pro you could write a SKILL.