---
title: "Demystifying ggplot2"
author: "Sara Burgess"
date: last-modified
format: 
    html: default
    gfm: default
knitr: 
    opts_chunk: 
      dev: "ragg_png"
---

Recently I was speaking to a PhD student about R. Our conversation went something like this (I'm paraphrasing):

> Student: "I hate R."
>
> Me: "Is it the modelling part you don't like?"
>
> Student: "All of it. R and I just don't see eye to eye."
>
> Me: "What about generating plots in R using ggplot2?"
>
> Student: "No, that takes so long, particularly adding in colours. I'd love to be able to generate beautiful plots though."
>
> Me: "It took me quite a while to understand how to use a custom palette and it's also quite fiddly getting the hang of all the different options you can add into the theme function to format a ggplot2 plot."

Hopefully you get the idea that she really doesn't like R. I felt for her, as writing up a PhD is hard enough without adding something else on top. This blog is for her, and for anyone else out there who feels like they'll never get their head around R and ggplot2. This was me a few years ago. I used to get frustrated with R and thought I'd never get the hang of it. I only started to get my head around it through practice, and it wasn't until ChatGPT emerged that I stopped feeling so frustrated. However, despite AI being a huge help with coding, it still helps to understand the basics and how to build up a plot. It wasn't until I began reading Nicola Rennie's blogs that I really appreciated the intricacies of plot design using ggplot2. It's not just about producing a beautiful plot; it's also about generating one that's easy for your audience to understand. In this blog I'll walk through how I build up a plot using ggplot2.

```{r}
#| echo: false
#| message: false
#| warning: false

#load libraries
#library(tidyverse)
library(gt)
#library(ggtext)
```

First load the libraries that we'll be using:

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

In this example we'll be generating a heatmap using a simulated dataset. This dataset gives the number of contigs carrying each antimicrobial resistance gene, broken down by bacterial family. The numbers are made up. You can load the data from here:

```{r}
#| message: false
#| warning: false
#| eval: false

df <- read_csv("https://raw.githubusercontent.com/sburgess2/blogs/main/2026/2026-08-05-demystifying-ggplot2/data/arg_contigs.csv")
```

This is what the first six lines of this data frame look like:

```{r}
#| echo: false
#| message: false
#| warning: false

#setwd(here::here("2026/2026-08-05-demystifying-ggplot2"))
df <- read_csv("data/arg_contigs.csv")
```

```{r}
head(df)
```

The first step is to tell R/ggplot2 what data you're using. Here, within the `ggplot()` function, we're telling R to use the data frame called `df`, that `gene` goes on the x-axis, `family` goes on the y-axis, and that `fill` specifies each tile will be coloured by `n_contigs`. When you run this you'll see the plot panel with the axes drawn but no data plotted.

```{r}
#| message: false
#| warning: false

ggplot(data = df, aes(x = gene, y = family, fill = n_contigs))
```

To actually plot the data we need to tell R what type of plot we want. We do this by adding a geom layer with `+`. There is a whole family of geoms, all starting with `geom_`, and if you start typing that in your editor you'll get a list of options. There are a lot of different geoms, plus many extensions you can use with ggplot2 (you can find a list [here](https://exts.ggplot2.tidyverse.org/gallery/)). We're generating a heatmap, so we'll use `geom_tile()`. Other packages provide their own heatmap functions too. Check out these blogs [here](https://r-graph-gallery.com/215-the-heatmap-function.html) and [here](https://www.datanovia.com/learn/machine-learning/clustering/heatmap) if you want to learn more about other heatmap functions.

```{r}
#| message: false
#| warning: false

ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile()
```

We now have a basic heatmap. Before we continue, let's have a look at some of the arguments you can use within `geom_tile()`. You can find out more about how to use a function and its arguments by typing `?` followed by the function name into the console, for example `?geom_tile`.

Generally, ggplot2 geoms share the same set of core arguments e.g. `mapping`, `data`, `show.legend`, `inherit.aes` and a few others. These are always listed in the help file (but sometimes they're tricky to follow). The `mapping` and `data` arguments are only needed if a layer is doing something different from the rest of the plot; otherwise they're specified once in `ggplot()`. Generally for a `geom_tile()` you won't need to change the defaults for these core arguments. However, you may want to use `inherit.aes = FALSE` if you don't want to inherit the plot data from `ggplot()` or use `show.legend = FALSE` if you do not want a legend.

The things that you are more likely to want to change in `geom_tile()` are passed through `...`, so they don't appear as named arguments, which is also why the help file can be confusing to read. For example, aesthetics such as those listed below.

```{r}
#| message: false
#| warning: false
#| echo: false

tribble(
  ~aesthetic, ~description,
  "fill", "Interior colour of the tile.",
  "colour", "Colour of the tile border. NA by default, so tiles are drawn without an outline.",
  "linewidth", "Thickness of the tile border. This has no visible effect until colour is set",
  "linetype", "Line style of the border, for example solid or dashed.",
  "alpha", "Opacity of the fill, from 0 (transparent) to 1 (opaque).",
  "width", "Width of each tile.",
  "height", "Height of each tile.") |>
     gt() |>
  cols_label(
    aesthetic = "Aesthetic",
    description = "Description"
  ) |>
  #tab_style(
   # style = cell_text(font = "monospace"),
    #locations = cells_body(columns = c(aesthetic, #description))
  #) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  cols_align(align = "left", columns = description) |> 
  tab_options(
    table.width = pct(100),
    table.font.size = px(14),
    data_row.padding = px(6),
    table.border.top.style = "none"
  )
```

Getting the hang of the different arguments is tricky. Part of it is practice, but part of it is accepting that you will need to look them up. I always do. Also ask yourself whether you really need to understand every argument. In this day and age of AI, do you really need to understand every line of code, especially if coding isn't your thing?

For this heatmap we're going to use the colour and linewidth aesthetics.

```{r}
#| message: false
#| warning: false

p <- ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile(colour = "white", linewidth = 1)

p
```

Next, we'll add the axis and legend labels. That's all you need if it's for a publication. For a presentation or a blog I would also add a title, which is set the same way using `labs(title = )`.

```{r}
#| message: false
#| warning: false

p_labels <- p +
    labs(x = "Antimicrobial resistance gene",
        y = "Family",
        fill = "Number of contigs"
    )

p_labels
```

The y-axis title, "Family" is currently rotated through 90 degrees and sitting halfway down the left-hand side. This is common practice for scientific publications but it requires the reader to tilt their head to read one word. I'd rather have it horizontal and above the y axis. There are arguments which allow the y axis title to be rotated and moved above the axis within the theme function. You can try it using the following code:

```{r}
#| message: false
#| warning: false

p_labels +
    theme(
        axis.title.y = element_text(
            angle = 0,
            vjust = 1,
            hjust = 1
        )
    )
```

`angle = 0` rotates the title back to horizontal, and `vjust = 1` moves it to the top of the axis. `hjust = 1` right-aligns it. However, when the y-axis title is set using `labs()` it always sits to the left. I would prefer it to be positioned directly above the family names. One option is to use `labs(tag = )`. A tag can be positioned by coordinates (within the theme function), so you can place it directly above the family names rather than out at the plot edge. We will use `tag` instead of `y`:

```{r}
#| message: false
#| warning: false

p_labels <- p +
    labs(
        x = "Antimicrobial resistance gene",
        y = NULL,
        fill = "Number of contigs",
        tag = "Family"
    )

p_labels
```

`y = NULL` drops the axis title and `tag = "Family"` puts the text back as a tag, which by default lands in the top left corner. The position of the tag is controlled by `plot.tag.position` in the `theme()` function. The `theme()` function will be the last layer we add to the plot (see below), which is when we'll reposition the tag. This tag sits on top of whatever is underneath, which means it will overlap your legend or your tiles if the coordinates are off.

Many figures in publications stop there, using the default formatting settings. However, to make your plot eye-catching and easy to read, choose colours wisely and declutter.

Let's change the colour palette to one that makes it easier to distinguish between the low and high values. You can either use a ready-made palette or create your own. The R Graph Gallery's [colour palette finder](https://r-graph-gallery.com/color-palette-finder) has a huge variety to choose from. If you find that list a bit overwhelming to look through, I'd recommend starting with the viridis palettes or the ColorBrewer palettes. For this example I'll use a ColorBrewer palette.

One thing to watch here: our tiles are coloured using `fill`, so we need a `scale_fill_*` function rather than `scale_colour_*`. Because `n_contigs` is a continuous variable we also need `scale_fill_distiller()`. `scale_fill_brewer()` looks like the obvious choice, but it only works with discrete variables. Within the `scale_fill_distiller()` argument we will use the arguments `palette` to specify which palette we want to use (you can find a list of them [here](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html)), and `direction` to specify the order where `direction = 1` is the default order of the palette and `direction = -1` reverses the order.

```{r}
#| message: false
#| warning: false

p_colour <- p_labels +
    scale_fill_distiller(palette = "YlGnBu", direction = 1)

p_colour
```

If you want to select your own colours, you can use `scale_fill_gradient()`, which takes a low colour and a high colour and works out everything in between. If you want a midpoint colour as well, `scale_fill_gradient2()` gives you low, mid and high.

```{r}
#| message: false
#| warning: false
#| eval: false

p_labels +
    scale_fill_gradient(low = "#F7FEAE", high = "#045275")

p_labels +
    scale_fill_gradient2(low = "#F7FEAE", mid = "#7CCBA2", high = "#045275")
```

Custom palettes for categorical fills work differently. If you want to use a custom palette, you first create the palette, one colour per category, then add it in using the `scale_fill_manual()` function:

```{r}
#| message: false
#| warning: false
#| eval: false

custom_colors <- c(
        "Category 1" = "#EDD746FF",
        "Category 2" = "#00496FFF",
        "Category 3" = "#DD4124FF",
        "Category 4" = "#B3B7B8FF")

ggplot(my_data, aes(x = category, y = value, fill = category)) +
    geom_col() +
    scale_fill_manual(values = custom_colors)
```

If you were colouring points or lines you'd use `scale_colour_manual()` instead.

Back to our heatmap. You'll notice that where `n_contigs = 0` the tile was coloured light yellow (as per the palette). However, you may want a value of zero to have either no colour or be set to the plot background colour. There are multiple ways to change this. The first option is to set limits within the `scale_fill_distiller()` function and assign white (or the background colour of your plot) to `na.value`.

```{r}
#| message: false
#| warning: false

p_colour <- p_labels +
    scale_fill_distiller(
        palette = "YlGnBu",
        direction = 1,
        limits = c(1, 10),
        breaks = c(2, 4, 6, 8, 10),
        na.value = "white"
    )

p_colour
```

Setting `limits = c(1, 10)` means the palette now starts at one rather than zero. Anything below the limits is treated as out of range and picks up `na.value`, so the zeros come out white. This is the version we'll carry forward. I have also included the argument `breaks` to change the numbers underneath the scale bar (if we didn't change these we'd get 2.5, 5.0, 7.5 and 10.0).

Another option is to convert the zeros to `NA` first then set the argument `na.value` to white (the default colour is `grey50`). If your dataset already has missing values instead of the zeros, you only need to add the `na.value` argument.

```{r}
#| message: false
#| warning: false
#| eval: false

#Convert zeros to NA
df <- df |>
    mutate(n_contigs = na_if(n_contigs, 0))

p <- ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile(colour = "white", linewidth = 1)

p_labels <- p +
    labs(x = "Antimicrobial resistance gene",
        y = NULL,
        fill = "Number of contigs",
        tag = "Family"
    )

p_labels +
    scale_fill_distiller(
        palette = "YlGnBu",
        direction = 1,
        na.value = "white"
    )
```

Alternatively you can use a binned scale. In other words, assign each value to a bin using the `scale_fill_stepsn()` function.

```{r}
#| message: false
#| warning: false
#| eval: false

p_labels +
    scale_fill_stepsn(
        colours = c("#EDF8B1", "#7FCDBB", "#225EA8", "#253494"),
        breaks = c(2, 4, 6, 8),
        limits = c(1, 10),
        na.value = "white"
    )
```

You may have noticed when using ggplot2 that the grey panel background is a ggplot2 default. We can see the outline of it around the heatmap. To remove this grey background we will use `theme_minimal()`. There are many other themes that you could try out that are also quite clean e.g. `theme_light()` and `theme_classic()`

```{r}
#| message: false
#| warning: false

p_theme <- p_colour + 
    theme_minimal(base_size = 12)

p_theme
```

Everything else can be formatted within the `theme()` function. Before the theme itself we will set the colours and fonts as objects at the top. It means that if you decide to change the font for the entire plot you can do this as a single edit instead of multiple edits.

Order matters here. `theme_minimal()` is a complete theme, which is why any additional formatting, including the tag position, needs to go into a single `theme()` call afterwards. If `theme()` is set first, it will be overridden by `theme_minimal()`.

```{r}
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 7

text_col <- "#1A1A1A"
label_col <- "#4D4D4D"
font <- "Times New Roman"
bg_col <- "white"

p_final <- p_theme +
    theme(
        legend.position = "top",
        legend.justification = "left",
        legend.location = "panel",
        legend.direction = "horizontal",
        legend.title.position = "top",
        legend.title = element_text(color = text_col, family = font),
        legend.text = element_text(color = label_col, family = font),
        plot.tag = element_text(
            hjust = 0,
            color = text_col, family = font),
        plot.tag.position = c(0.05, 0.87),
        plot.tag.location = "plot",
        plot.margin = margin(10, 10, 10, 10),
        axis.text.y = element_text(colour = label_col, family = font, size = rel(0.8)),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_text(colour = label_col, family = font),
        panel.background = element_rect(fill = bg_col, color = bg_col),
        plot.background = element_rect(fill = bg_col, color = bg_col)
    )

p_final
```

Here are some descriptions of what's been included in the `theme()` function:

- `legend.position = "top"` puts the key at the top of the plot.
- `legend.justification = "left"` left aligns it instead of centring it.
- `legend.location = "panel"` lines it up with the left edge of the tiles. Using `"plot"` would line it up with the far left edge of the whole plot, where the family names and the tag sit.
- `legend.direction = "horizontal"` lays the colour bar out sideways, which suits a wide heatmap because it doesn't eat width the tiles could be using.
- `legend.title.position = "top"` moves "Number of contigs" above the colour bar rather than leaving it alongside.
- `legend.title` and `legend.text` set the font and colour of the legend title and the numbers underneath it, using `element_text()`.
- `plot.tag` sets the font and colour of the tag, which is standing in for our y-axis title.
- `plot.tag.position = c(0.05, 0.87)` places the tag 5% in from the left and 87% of the way up. Both numbers need nudging by eye, and they'll need revisiting if you change the figure dimensions.
- `plot.tag.location = "plot"` measures those coordinates against the whole plot rather than the panel.
- `plot.margin = margin(10, 10, 10, 10)` sets the space around the outside of the plot, clockwise from the top.
- `axis.text` is the tick labels, so the gene names and the family names. Setting it here covers both axes at once.
- `axis.text.x = element_text(angle = 45, hjust = 1)` angles the gene names at 45 degrees. Generally I prefer my x axis labels to be horizontal, but some of the gene names are quite long and won't easily fit. `hjust = 1` keeps the end of each label next to the tile it belongs to.
- `axis.title.x` is the name of the axis itself, which is a separate element from `axis.text`. We only need the `.x` version because the y-axis title was dropped when we swapped it for the tag, so there's no `axis.title.y` left to style.
- `panel.background` and `plot.background` set the fill behind the tiles and behind the whole plot. Both use `element_rect()` rather than `element_text()` because they're rectangles rather than text.

One last thing: the plot in the plot pane is drawn at whatever size the pane happens to be. Drag the pane wider and the text gets relatively smaller, make it narrower and your labels start colliding. You end up tuning text sizes against a moving target.

To view your plot at the same dimensions it will be exported at, use the camcorder package. `gg_record()` sets a fixed canvas, and from then on your plot is generated in the viewer (instead of the plot window) at those exact dimensions:

```{r}
#| eval: false

library(camcorder)

gg_record(
    dir = "output/recording",
    device = "png",
    width = 12,
    height = 7,
    units = "in",
    dpi = 300
)

ggplot(data = df, aes(x = gene, y = family, fill = n_contigs)) +
    geom_tile(colour = "white", linewidth = 1) +
    labs(
        x = "Antimicrobial resistance gene",
        y = NULL,
        fill = "Number of contigs",
        tag = "Family"
    ) +
    scale_fill_distiller(
        palette = "YlGnBu",
        direction = 1,
        limits = c(1, 10),
        breaks = c(2, 4, 6, 8, 10),
        na.value = "white"
    ) +
    theme_minimal(base_size = 12) +
    theme(
        legend.position = "top",
        legend.justification = "left",
        legend.location = "panel",
        legend.direction = "horizontal",
        legend.title.position = "top",
        legend.title = element_text(color = text_col, family = font),
        legend.text = element_text(color = label_col, family = font),
        plot.tag = element_text(
            hjust = 0, 
            color = text_col, family = font),
        plot.tag.position = c(0.05, 0.87),
        plot.tag.location = "plot",
        plot.margin = margin(10, 10, 10, 10),
        axis.text = element_text(colour = label_col, family = font, size = rel(0.8)),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_text(colour = label_col, family = font),
        panel.background = element_rect(fill = bg_col, color = bg_col),
        plot.background = element_rect(fill = bg_col, color = bg_col)
    )

gg_stop_recording()
```

To save the file use `ggsave()`:

```{r}
#| eval: false

ggsave(
    "output/arg_heatmap.png",
    #p_final,
    width = 12,
    height = 7,
    dpi = 300,
    bg = bg_col
)
```

You may be asking whether it's a nuisance writing out all that `theme()` code every time. It is, and there are a few ways to avoid it:

- Write a script template, as described by [Nicola Rennie](https://nrennie.rbind.io/blog/script-templates-r/).
- Build your own custom theme, described [here](https://rfortherestofus.com/2025/04/ggplot2-theme) by R for the Rest of Us.
- If you use Claude, write the whole thing up as a skill so it applies your house style for you.

Any of those turns the theme block into something you write once rather than every time.

So that's how I build up a plot. Most of those lines had nothing to do with the data, which is what I'd want my PhD student to take away. Getting a heatmap on screen takes two lines. The formatting afterwards is where the afternoon goes.