---
title: "Adding annotations"
author: "Sara Burgess"
format: html
editor: visual
---

## Adding annotations to your plots

I think one of the trickiest thing about puttig together a data visualisation in R is annotating the plot. Annotations can really help your reader interpret the graph / take off some of the cognitive load, highlight a certain observation or group of observations. There are multiple ways to do this. Using the following functions:



```{r}
#| warning: false
#| message: false
#| echo: false

library(tidyverse)

tuesdata <- tidytuesdayR::tt_load("2026-02-03")

edible_plants <- tuesdata$edible_plants
```


```{r}
#| warning: false
#| message: false
#| echo: false

edible_tidy <- edible_plants %>%

  select(
    taxonomic_name,
    common_name,
    cultivation,
    temperature_class,

    temperature_growing,
    days_harvest
  ) %>%

  filter(
    !cultivation == "Miscellaneous",

    !days_harvest == "continual"
  ) %>%

  drop_na() %>%

  separate(
    temperature_growing,

    into = c("temp_min", "temp_max"),

    sep = "-",
    fill = "right",
    remove = FALSE
  ) %>%

  separate(
    days_harvest,

    into = c("days_min", "days_max"),

    sep = "-",
    fill = "right",
    remove = FALSE
  )

write_csv("blogs/2026/2026-08-24-annotations/data/edible_plants_tidy.csv")
```

Lets use a TidyTuesday dataset (from the ...) to illustrate how these different functions work. 

First load the libraries that we'll be using:

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

I've generated a tidy version of the edible plants dataset (the orignal dataset can be found here). You can load the tidy dataset from here:

```{r}
#| message: false
#| warning: false
#| eval: false

df <- read_csv(
  "https://raw.githubusercontent.com/sburgess2/blogs/main/2026/2026-08-24-annotations/data/edible_plants_tidy.csv"
)
```

First we'll generate a a basic bar plot. 
```{r}
#| warning: false
#| message: false

ggplot(
  df_all,

  aes(
    x = factor(common_name, levels = veg_levels_rev),

    y = days_min
  )
) +

  geom_col(fill = light_gr) +

  geom_text(aes(label = days_min), hjust = 1.5, size = 2, color = "#ffffff") +

  theme_minimal(base_family = font, base_size = 7)
```

```{r}
#| warning: false
#| message: false

```

Using coord_cartesian(clip = "off") allows annotations to be drawn outside of the plot. However, when using geom_text with treemaps if the coordinates are outside the plot it will expand the treemap to fit. Therefore for annotations outside the use geom_textbox where each annotation uses a separate dataframe. See treemap.R and 03_plotting.R for examples.

geom_textbox and annotate() use mm as the size by default. 10pt = 3.528 mm. They can not use unit.size = "pt" to change to put

```         
size = 3.528, #annotate doesn't have an option to change to pt. 
```

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).