---
title: "Adding annotations to your plots"
author: "Sara Burgess"
format: html
editor: visual
---

I think one of the trickiest things about generating a plot in R is annotating it. Annotations can really help your reader interpret your graph and highlight a certain observation or group of observations. As with many things in R there are multiple ways to do this. In this blog I'll run through six functions that can be used for annotating.

```{r}
#| warning: false
#| message: false
#| echo: false

library(tidyverse)

tuesdata <- tidytuesdayR::tt_load("2026-02-03")

edible_plants <- tuesdata$edible_plants
```

```{r}
#| warning: false
#| message: false
#| echo: false

edible_tidy <- edible_plants %>%

  select(
    taxonomic_name,
    common_name,
    cultivation,
    temperature_class,

    temperature_growing,
    days_harvest
  ) %>%

  filter(
    !cultivation == "Miscellaneous",

    !days_harvest == "continual"
  ) %>%

  drop_na() %>%

  separate(
    temperature_growing,

    into = c("temp_min", "temp_max"),

    sep = "-",
    fill = "right",
    remove = FALSE
  ) %>%

  separate(
    days_harvest,

    into = c("days_min", "days_max"),

    sep = "-",
    fill = "right",
    remove = FALSE
  ) |>
  mutate(across(ends_with(c("_min", "_max")), as.numeric)) |>
  mutate(
    common_name = recode(
      common_name,

      "Pea" = "Peas",

      "Beans (Runner)" = "Beans climbing/snake",

      "Beans (French)" = "Beans dwarf/bush",

      "Beans (Broad)" = "Broad bean",

      "Cabbage (Spring)" = "Cabbage loose-headed",

      "Zucchini or Courgette" = "Zucchini",

      "Cucumber" = "Cucumber",

      "Melon" = "Rockmelon",

      "Water melon" = "Watermelon",

      "Lettuce (Headed)" = "Lettuce",

      "Artichoke (Globe)" = "Artichokes globe",

      "Aubergine" = "Aubergine",

      "Beans (Borlotti)" = "Beans dwarf/bush",

      "Beetroot" = "Beetroot",

      "Bok Choy" = "Bok Choy",

      "Cabbage (Autumn red)" = "Cabbage tight-headed",

      "Cabbage (Autumn)" = "Cabbage tight-headed",

      "Cabbage (Chinese)" = "Chinese cabbage",

      "Cabbage (Spring red)" = "Cabbage loose-headed",

      "Cabbage (Summer red)" = "Cabbage tight-headed",

      "Cabbage (Summer)" = "Cabbage tight-headed",

      "Chilli Pepper" = "Chilli peppers",

      "Fennel" = "Fennel",

      "Mangetout" = "Mangetout",

      "Onion" = "Onion",

      "Parsnip" = "Parsnip",

      "Bell Pepper" = "Bell peppers",

      "Potatoes (Main crop)" = "Potato",

      "Squash (Summer)" = "Summer squash",

      "Squash (Winter)" = "Squash",

      "Tomato (Regular)" = "Tomato"
    )
  ) %>%

  distinct() %>%

  filter(
    !common_name == "Cabbage loose-headed" &
      !cultivation == "Legume" &
      !common_name == "Brussels Sprouts" &
      !common_name == "Potatoes (Early)" &
      !common_name == "Tomato (Cherry)"
  ) |>
  mutate(
    taxonomic_name = str_squish(str_remove(taxonomic_name, "\\(.*\\)")),
    taxonomic_name = str_to_sentence(taxonomic_name)
  )


write_csv(
  edible_tidy,
  "2026/2026-08-24-annotations/data/edible_plants_tidy.csv"
)
```

```{r}
#| echo: false
#| message: false
#| warning: false

#To run the line below directly in the console use
#setwd(here::here("2026", "2026-08-24-annotations"))
edible_tidy <- read_csv(
  "2026/2026-08-24-annotations/data/edible_plants_tidy.csv"
)
```

We will use the [Edible Plants](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-02-03/readme.md) TidyTuesday dataset (from the [Data Science Learning Community](https://dslc.io/)) to illustrate how these different functions work.

First load the libraries that we'll be using:

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(ggrepel)
library(ggtext)
library(directlabels)
```

I've generated a tidy version of the Edible Plants dataset. You can load the tidy dataset from here:

```{r}
#| message: false
#| warning: false
#| eval: false

edible_tidy <- read_csv(
  "https://raw.githubusercontent.com/sburgess2/blogs/main/2026/2026-08-24-annotations/data/edible_plants_tidy.csv"
)
```

First we'll generate a bar plot.

```{r}
#| warning: false
#| message: false

p <- ggplot(
  data = edible_tidy,
  aes(
    y = fct_reorder(common_name, days_min), #Reorder common_name by days_min
    x = days_min
  )
) +
  geom_col(fill = "#55692f") +
  labs(
    y = NULL,
    subtitle = "Common name",
    x = "Minimum number of days to harvest"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot", #align subtile to the plot
    panel.grid = element_blank() #remove gridlines
  )
```

## Annotating every row of your data using `geom_text()`

For this first example of annotating a plot we'll use `geom_text()` to label each bar with its minimum number of days to harvest.

```{r}
#| warning: false
#| message: false

p_label <- p +
  geom_text(
    aes(label = days_min),
    size = 10 / .pt,
    hjust = 1,
    nudge_x = -1,
    colour = "#ffffff"
  )
p_label
```

In this example, we've used `size` to specify the size of the text in mm, `hjust` to align the text to the right (by default this is set to 0.5, which centres the text; 0 left aligns it and 1 right aligns it), `nudge_x` to move the label to the left slightly so it's not right up against the edge of the bar, and `colour` to colour the text white.

Annotations can also be a great way to highlight particular observations. Let's use lettuce as an example. First we'll try adding one annotation using `geom_text()`.

```{r}
#| warning: false
#| message: false

p +
  geom_text(
    y = "Lettuce",
    x = 45,
    label = "Lettuce has a wide temperature growth range",
    size = 10 / .pt,
    hjust = 0
  )
```

Perhaps you may have noticed that the annotation looks a bit fuzzy. This is because when you use `geom_text()` it's inheriting the data from the dataframe `edible_tidy`. There are 30 observations, so the text `Lettuce has a wide temperature growth range` gets placed onto the plot 30 times. We will come back to `geom_text()`, but for now let's look at another way to place this annotation using `annotate()`.

## Single annotations using `annotate()`

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Lettuce",
    x = 45,
    label = "Lettuce has a wide temperature growth range",
    size = 10 / .pt,
    hjust = 0
  )
```

Now our label looks cleaner. Let's look at some of the arguments we've used. First we've specified the type of geom with `geom = "text"`. In this example we want to add a text annotation, but `annotate()` can also be used for adding shapes and lines (see the example below). `x` and `y` are used for specifying the position of the annotation, again `size` is used (`.pt` is a constant ggplot2 exports, so `10 / .pt` gives you 10pt text) and `hjust` is used to align the text to the left.

Next we'll look at how `annotate()` can be used to add a line or an arrow.

A line:

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Lettuce",
    x = 50,
    label = "Lettuce has a wide temperature growth range",
    size = 10 / .pt,
    hjust = 0
  ) +
  annotate(
    geom = "segment",
    x = 50,
    y = "Lettuce",
    xend = 45,
    yend = "Lettuce"
  )
```

An arrow:

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Lettuce",
    x = 50,
    label = "Lettuce has a wide temperature growth range",
    size = 10 / .pt,
    hjust = 0
  ) +
  annotate(
    geom = "segment",
    x = 50,
    y = "Lettuce",
    xend = 45,
    yend = "Lettuce",
    arrow = arrow(
      length = unit(0.2, "cm"), #specifies the length of the arrowhead
      type = "closed" #specifies the type of arrowhead
    )
  )
```

In the examples above `x` and `y` are used to specify the coordinates of the start of the line or arrow and `xend` and `yend` the end of the line or arrow. You could add a line with `geom_segment()`, but it has the same problem as `geom_text()`: you get one segment per row of the data, all stacked in the same place. The alternatives are `annotate(geom = "segment", ...)` as above, or passing `geom_segment()` its own one-row data frame.

Other aesthetics can also be included as arguments in `annotate()`, for example the colour or transparency of the text or arrow.

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Lettuce",
    x = 50,
    label = "Lettuce has a wide temperature growth range",
    size = 10 / .pt,
    hjust = 0,
    colour = "blue"
  ) +
  annotate(
    geom = "segment",
    x = 50,
    y = "Lettuce",
    xend = 45,
    yend = "Lettuce",
    alpha = 0.5,
    arrow = arrow(
      length = unit(0.2, "cm"), #specifies the length of the arrowhead
      type = "closed" #specifies the type of arrowhead
    )
  )
```

You can also italicise part of the label by writing it as a plotmath expression and setting `parse = TRUE`:

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Watermelon",
    x = 80,
    label = "paste(italic('Phytophthora infestans'), ' can cause blight in potatoes and tomatoes')",
    parse = TRUE,
    size = 10 / .pt,
    hjust = 0,
    colour = "blue"
  ) +
  annotate(
    geom = "curve",
    x = 100,
    y = 23 + 0.3,
    xend = 100,
    yend = "Potato",
    curvature = 0.3,
    #angle = 90,
    #ncp = 10,
    alpha = 0.5,
    arrow = arrow(
      length = unit(0.2, "cm"),
      type = "closed"
    )
  ) +
  annotate(
    geom = "curve",
    x = 100,
    y = 23 - 0.3,
    xend = 65,
    yend = "Tomato",
    curvature = -0.3,
    #angle = 90,
    alpha = 0.5,
    arrow = arrow(
      length = unit(0.2, "cm"),
      type = "closed"
    )
  )
```

Generally `annotate()` is fine for most annotations, but there are a few things it won't do:

Formatting is limited to plotmath. The *Phytophthora infestans* label above works, but it's quite a long expression. For one italic word that's fine, but for multiple font face changes the code becomes very wordy.

It won't wrap text. As you may have noticed in the above example, the annotation is drawn on a single line, so a long sentence runs off the edge of the panel. You can break the line yourself with `\n`, but then every break has to be checked again whenever the plot changes size. `\n` also doesn't work with plotmath.

You cannot style a box separately from the text. `geom = "label"` will draw one, but its border colour comes from `colour`, which also sets the text colour.

## Annotations with `geom_richtext()` and `geom_textbox()`

The {ggtext} package gives you two geoms that can use Markdown and HTML.

`geom_richtext()` is the one to reach for when the annotation is short but part of it needs different styling: an italic species name, a coloured word to match a colour in the plot, a superscript unit.

When using `geom_richtext()` or `geom_textbox()`, each annotation gets its own data frame and `inherit.aes = FALSE` so that it ignores the mapping set in `ggplot()`.

```{r}
#| warning: false
#| message: false

blight <- tibble(
  x = 1,
  y = "Potato",
  label = "*Phytophthora infestans* causes blight in potatoes"
)

p +
  geom_richtext(
    data = blight,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 10 / .pt,
    hjust = 0,
    fill = NA,
    label.colour = NA,
    colour = "white"
  )
```

Above we've used `fill = NA` and `label.colour = NA` to strip off the background and border that `geom_richtext()` draws by default. In the example below the annotation sits in a coloured box, and because `geom_richtext()` doesn't wrap text we've used `<br>` to break up the sentence.

```{r}
#| warning: false
#| message: false

blight_textbox <- tibble(
  x = 80,
  y = "Watermelon",
  label = "*Phytophthora infestans*<br>causes blight in <br>potatoes and tomatoes"
)

p +
  geom_richtext(
    data = blight_textbox,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 10 / .pt,
    hjust = 0,
    fill = "#99A582",
    label.colour = "#55692f",
    colour = "#2b3418"
  ) +
  annotate(
    geom = "curve",
    x = 100,
    y = 22 + 0.5,
    xend = 100,
    yend = "Potato",
    curvature = 0.3,
    alpha = 0.5,
    arrow = arrow(
      length = unit(0.2, "cm"),
      type = "closed"
    )
  ) +
  annotate(
    geom = "curve",
    x = 100,
    y = 22 - 0.5,
    xend = 65,
    yend = "Tomato",
    curvature = -0.3,
    alpha = 0.5,
    arrow = arrow(
      length = unit(0.2, "cm"),
      type = "closed"
    )
  )
```

I'd recommend using `geom_textbox()` when the annotation is long enough to need wrapping. You can specify a `width` and it wraps the text to fit.

```{r}
#| warning: false
#| message: false

blight_wrapped <- tibble(
  x = 80,
  y = "Watermelon",
  label = "*Phytophthora infestans* causes blight in potatoes and tomatoes"
)

p +
  geom_textbox(
    data = blight_wrapped,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 10 / .pt,
    width = unit(5, "cm"),
    hjust = 0,
    box.colour = "grey80",
    fill = "grey98"
  )
```

The thing that used to catch me out was the difference between `hjust` and `halign`. `hjust` positions the whole box relative to `x` and `halign` aligns the text inside the box.

One thing `geom_textbox()` won't do is rotate the text to an arbitrary angle. It has an `orientation` argument that turns the box in 90° steps and nothing in between. If you need text on a diagonal, `geom_richtext()` takes an `angle` aesthetic.

## Labelling groups directly with `geom_dl()`

One function I hadn't used before writing this blog was `geom_dl()`. It comes from the {directlabels} package and is generally used for labelling data directly rather than using a legend. The `method` argument is used for specifying how the labels will be placed. For example, `method = "last.points"` puts one label at the end of each line, which works well for line graphs. In our example we will use `method = "smart.grid"` to put one label per group in the middle of that group's cloud of points.

```{r}
#| warning: false
#| message: false

ggplot(
  data = edible_tidy,
  aes(
    x = temp_min,
    y = days_min,
    colour = cultivation
  )
) +
  geom_point(size = 2, alpha = 0.8) +
  geom_dl(
    aes(label = cultivation),
    method = "smart.grid"
  ) +
  scale_colour_brewer(palette = "Dark2") +
  guides(colour = "none") +
  labs(
    x = "Minimum growing temperature (°C)",
    y = "Minimum days to harvest"
  ) +
  theme_minimal()
```

## Getting the annotation in the right place

This is the part I find hardest, and there isn't one trick that fixes it. Even asking Claude or another AI tool tends to produce coordinates that are roughly right and then need nudging anyway. Here are a few suggestions to help:

It helps to anchor the coordinate to the data rather than typing a number. A number is correct for exactly the data you had in front of you when you typed it, whereas something like `x = max(edible_tidy$days_min)` still works after you add rows or swap the dataset.

`I()` can be used to switch from data units to a proportion of the panel, so `x = I(0.05), y = I(0.95)` means 5% across and 95% up.

On a discrete axis you can pass either the level name (`y = "Lettuce"`) or a number. The numbers are the level positions, so `y = 12.5` sits halfway between the twelfth and thirteenth bar.

Then there's output size, which you should decide before you position anything. Text size is absolute, so 10pt is 10pt, while the panel stretches to fill whatever output size it is drawn on. An annotation that sits neatly in the plot pane can end up across a bar in the 8×6 inch PNG you export. I use the {camcorder} package to render my plots at the dimensions I export them in:

```{r}
#| eval: false

camcorder::gg_record(
  dir = "plots",
  width = 8,
  height = 6,
  units = "in",
  dpi = 300
)
```

The {ggannotate} package is also a great way to get an estimate of the coordinates. It launches a Shiny app in which you click on the plot where you want the annotation, and it hands back the code with the coordinates filled in. Install it with `remotes::install_github("MattCowgill/ggannotate")` and call `ggannotate(p)`.

Often you will find that you will need to nudge the coordinates slightly to get them in the correct position.

### Annotations outside the panel

Another function I wish I'd learnt about when I first started using ggplot2 is `coord_cartesian(clip = "off")`. It lets annotations be drawn outside the plot. ggplot2 clips anything beyond the panel edges by default, so turning clipping off is what allows the text into the margin, and then the margin has to be wide enough to hold it.

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    x = max(edible_tidy$days_min) * 1.02,
    y = "Lettuce",
    label = "Days to harvest depend\non when the crop is sown",
    size = 10 / .pt,
    hjust = 0,
    colour = "grey40"
  ) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5, 60, 5, 5))
```

`margin()` takes its values in the order top, right, bottom, left, and the units are points, so 60 on the right leaves roughly 2 cm for the label to sit in.

## Repelling overlapping labels with `geom_text_repel()`

`geom_text_repel()` is generally used when you have overlapping labels. As the name suggests, the labels repel each other so that they are placed in free space rather than overlapping. A scatterplot is a better example of this.

```{r}
#| warning: false
#| message: false

p <- ggplot(
  data = edible_tidy,
  aes(
    y = days_min,
    x = temp_min,
    label = common_name
  )
) +
  geom_point(pch = 21, fill = "#55692f") +
  labs(
    y = NULL,
    subtitle = "Minimum number of days to harvest",
    x = "Minimum growing temperature (°C)"
  ) +
  geom_text_repel(
    size = 5,
    box.padding = 0.5
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot" #align subtile to the plot
    #panel.grid = element_blank() #remove gridlines
  )

p
```

For some other great examples of where `geom_text_repel()` can be used, see [this post](https://r-graph-gallery.com/package/ggrepel.html) on the r-graph-gallery.

## A note on font size

Almost every geom here measures `size` in millimetres, while the rest of ggplot2 (axis text, titles, theme elements) uses points.

`geom_text()` gained a `size.unit` argument in ggplot2 3.5.0, so you can write `size = 10, size.unit = "pt"` and be done. It doesn't help with `annotate()`, {ggrepel} or {ggtext}. The version that works everywhere is `.pt`, a constant ggplot2 exports for exactly this conversion:

```{r}
#| warning: false
#| message: false

ggplot() +
  annotate(
    geom = "text",
    x = 0,
    y = 2,
    label = "size = 10",
    size = 10
  ) +
  annotate(
    geom = "text",
    x = 0,
    y = 1,
    label = "size = 10 / .pt",
    size = 10 / .pt
  ) +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(0, 3)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10)
```

The point size you want, divided by `.pt`, gives the millimetre value that produces it.

## Summary

| Function | Use | Aesthetics | Word wrap | Text formatting | Text rotation | Box outline | Font size units | Package |
|--------|--------|--------|--------|--------|--------|--------|--------|--------|
| `geom_text()` | Label every row of the data | Uses `aes()`, inherits from `ggplot()` | No — break lines with `\n` | Plain text, or plotmath with `parse = TRUE` | Yes (`angle`) | No | mm, or `size.unit = "pt"` | ggplot2 |
| `annotate()` | Single annotations not tied to a row of data | No `aes()`, values passed straight in | No | Plain text or plotmath | Yes (`angle`) | Only via `geom = "label"` | mm — use `10 / .pt` | ggplot2 |
| `geom_text_repel()` | Many labels that would otherwise overlap | Uses `aes()`, inherits from `ggplot()` | No | Plain text or plotmath | Yes (`angle`) | No | mm | ggrepel |
| `geom_richtext()` | Short labels needing inline styling | Uses `aes()`, needs its own data frame | No | Markdown / HTML subset | Yes (`angle`) | Yes (`label.colour`, `label.size`) | mm | ggtext |
| `geom_textbox()` | Sentences and paragraphs | Uses `aes()`, needs its own data frame | Yes (`width`) | Markdown / HTML subset | 90° steps only (`orientation`) | Yes (`box.colour`, `box.size`) | mm | ggtext |
| `geom_dl()` | Labelling groups directly instead of using a legend | Uses `aes()` plus a `method` | No | Plain text | Via the method (e.g. `rot = 30`) | Via the method (e.g. `angled.boxes`) | `cex` multiplier | directlabels |