---
title: "Adding annotations"
author: "Sara Burgess"
format: html
editor: visual
---

## Adding annotations to your plots

I think one of the trickiest thing about puttig together a data visualisation in R is annotating the plot. Annotations can really help your reader interpret the graph / take off some of the cognitive load, highlight a certain observation or group of observations. There are multiple ways to do this. Using the following functions:



```{r}
#| warning: false
#| message: false
#| echo: false

library(tidyverse)

tuesdata <- tidytuesdayR::tt_load("2026-02-03")

edible_plants <- tuesdata$edible_plants
```


```{r}
#| warning: false
#| message: false
#| echo: false

edible_tidy <- edible_plants %>%

  select(
    taxonomic_name,
    common_name,
    cultivation,
    temperature_class,

    temperature_growing,
    days_harvest
  ) %>%

  filter(
    !cultivation == "Miscellaneous",

    !days_harvest == "continual"
  ) %>%

  drop_na() %>%

  separate(
    temperature_growing,

    into = c("temp_min", "temp_max"),

    sep = "-",
    fill = "right",
    remove = FALSE
  ) %>%

  separate(
    days_harvest,

    into = c("days_min", "days_max"),

    sep = "-",
    fill = "right",
    remove = FALSE
  ) |>
  mutate(across(ends_with(c("_min", "_max")), as.numeric)) |>
  mutate(
    common_name = recode(
      common_name,

      "Pea" = "Peas",

      "Beans (Runner)" = "Beans climbing/snake",

      "Beans (French)" = "Beans dwarf/bush",

      "Beans (Broad)" = "Broad bean",

      "Cabbage (Spring)" = "Cabbage loose-headed",

      "Zucchini or Courgette" = "Zucchini",

      "Cucumber" = "Cucumber",

      "Melon" = "Rockmelon",

      "Water melon" = "Watermelon",

      "Lettuce (Headed)" = "Lettuce",

      "Artichoke (Globe)" = "Artichokes globe",

      "Aubergine" = "Aubergine",

      "Beans (Borlotti)" = "Beans dwarf/bush",

      "Beetroot" = "Beetroot",

      "Bok Choy" = "Bok Choy",

      "Cabbage (Autumn red)" = "Cabbage tight-headed",

      "Cabbage (Autumn)" = "Cabbage tight-headed",

      "Cabbage (Chinese)" = "Chinese cabbage",

      "Cabbage (Spring red)" = "Cabbage loose-headed",

      "Cabbage (Summer red)" = "Cabbage tight-headed",

      "Cabbage (Summer)" = "Cabbage tight-headed",

      "Chilli Pepper" = "Chilli peppers",

      "Fennel" = "Fennel",

      "Mangetout" = "Mangetout",

      "Onion" = "Onion",

      "Parsnip" = "Parsnip",

      "Bell Pepper" = "Bell peppers",

      "Potatoes (Main crop)" = "Potato",

      "Squash (Summer)" = "Summer squash",

      "Squash (Winter)" = "Squash",

      "Tomato (Regular)" = "Tomato"
    )
  ) %>%

  distinct() %>%

  filter(
    !common_name == "Cabbage loose-headed" &
      !cultivation == "Legume" &
      !common_name == "Brussels Sprouts"
  )

write_csv(
  edible_tidy,
  "2026/2026-08-24-annotations/data/edible_plants_tidy.csv"
)
```

```{r}
#| echo: false
#| message: false
#| warning: false

#To run the line below directly in the console use
#setwd(here::here("2026", "2026-08-24-annotations"))
edible_tidy <- read_csv(
  "2026/2026-08-24-annotations/data/edible_plants_tidy.csv"
)
```

Lets use a TidyTuesday dataset (from the ...) to illustrate how these different functions work. 

First load the libraries that we'll be using:

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

I've generated a tidy version of the edible plants dataset (the orignal dataset can be found here). You can load the tidy dataset from here:

```{r}
#| message: false
#| warning: false
#| eval: false

edible_tidy <- read_csv(
  "https://raw.githubusercontent.com/sburgess2/blogs/main/2026/2026-08-24-annotations/data/edible_plants_tidy.csv"
)
```

First we'll generate a bar plot. 
```{r}
#| warning: false
#| message: false

p <- ggplot(
  data = edible_tidy,
  aes(
    y = fct_reorder(common_name, days_min), #Reorder common_name by days_min
    x = days_min
  )
) +
  geom_col(fill = "#ddd69e") +
  labs(
    y = NULL,
    subtitle = "Minimum number of days to harvest",
    x = "Common name"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot", #align subtile to the plot
    panel.grid = element_blank() #remove gridlines
  )
```

Annotations can be a great way to highlight particular observations. Lets use lettuce as an example. First we'll try adding one annotation using `geom_text()`.
```{r}
#| warning: false
#| message: false

p +
  geom_text(
    y = "Lettuce",
    x = 45,
    label = "Lettuce has a wide temperature growth range",
    size = 4.23,
    hjust = 0
  )
```

Perhaps you may have noticed that the annotation looks a bit fuzzy. This is because when you use `geom_text()` it's inheriting the data from the dataframe `edible_tidy`. There are 32 observations, so the text `Lettuce has a wide temperature growth range` gets placed onto the plot 32 times. We will come back to geom_text but for now lets look at another way to place to this annotation using `annotate()`

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Lettuce",
    x = 45,
    label = "Lettuce has a wide temperature growth range",
    size = 3.53,
    hjust = 0
  )
```

Now our label is cleaning looking. Lets look at some of the arguments we've used. First we've specified the type of geom to use. In this case we want to add a text annotation, but `annotate()` can also be used for adding shapes and lines. `x` and `y` are used for specifying the position of the annotation, the size of the text is specified in mm. 10pt is 3.53 mm. `hjust` is used to horizontally align the text. By default this is set to 0.5, which centres the text. 0 left aligns the text and 1 right alighns the text.

Next lets look at how we can use annotate() to add a line or an arrow. 
```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Lettuce",
    x = 60,
    label = "Lettuce has a wide temperature growth range",
    size = 3.53,
    hjust = 0
  ) +
  annotate(
    geom = "segment",
    x = 60,
    y = "Lettuce",
    xend = 45,
    yend = "Lettuce"
  )
```

```{r}
#| warning: false
#| message: false

p +
  annotate(
    geom = "text",
    y = "Lettuce",
    x = 60,
    label = "Lettuce has a wide temperature growth range",
    size = 3.53,
    hjust = 0
  ) +
  annotate(
    geom = "segment",
    x = 60,
    y = "Lettuce",
    xend = 45,
    yend = "Lettuce",
    arrow = arrow(
      length = unit(0.2, "cm"), #specifies the length of the arrowhead
      type = "closed" #specifies the type of arrowhead
    )
  )
```

Note to self don't use geom_segment() as again it will plot one segment per observation unless you use a separate df. 
```{r}
#| warning: false
#| message: false

```

Using coord_cartesian(clip = "off") allows annotations to be drawn outside of the plot. However, when using geom_text with treemaps if the coordinates are outside the plot it will expand the treemap to fit. Therefore for annotations outside the use geom_textbox where each annotation uses a separate dataframe. See treemap.R and 03_plotting.R for examples.

geom_textbox and annotate() use mm as the size by default. 10pt = 3.528 mm. They can not use unit.size = "pt" to change to put

```         
size = 3.528, #annotate doesn't have an option to change to pt. 
```

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).